---
title: "Access and analyse EOPF STAC Zarr data with R"
format:
  html:
    df-print: tibble
execute:
  cache: true
---

### Introduction

In this tutorial, we will demonstrate how to access EOPF Zarr products directly from the [EOPF Sentinel Zarr Sample Service STAC Catalog](https://stac.browser.user.eopf.eodc.eu/) using R. We will introduce R packages that enable us to effectively get an overview of and read Zarr arrays.

### What we will learn

- â˜ï¸ How to overview cloud-optimised datasets from the EOPF Zarr STAC Catalog with the `Rarr` package.
- ðŸ”Ž Read and examine loaded datasets
- ðŸ“Š Perform simple data analyses with the loaded data

### Prerequisites

An R environment is required to follow this tutorial, with R version >= 4.4.0. For local development, we recommend using either [RStudio](https://p!osit.co/download/rstudio-desktop/) or [Positron](https://posit.co/products/ide/positron/) and making use of [RStudio projects](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) for a self-contained coding environment.

We will use the following packages in this tutorial:
- [`rstac`](https://brazil-data-cube.github.io/rstac/) (for accessing the STAC catalog)
- [`tidyverse`](https://tidyverse.tidyverse.org/) (for data manipulation)
- [`stars`](https://r-spatial.github.io/stars/)) (for working with spatiotemporal data)
- [`terra`](https://rspatial.github.io/terra/index.html) (for working with spatial data in raster format).

You can install them directly from CRAN:

```{r install}
#| eval: false
install.packages("rstac")
install.packages("tidyverse")
install.packages("stars")
install.packages("terra")
```

We will also use the `Rarr` package to read Zarr data. It must be installed from Bioconductor, so first install the `BiocManager` package:

```{r install-BiocManager}
#| eval: false
install.packages("BiocManager")
```

Then, use this package to install `Rarr`:

```{r install-rarr}
#| eval: false
BiocManager::install("Rarr")
```

Finally, load the packages into your environment:

```{r load, message=FALSE, warning=FALSE}
#| cache: false
library(rstac)
library(tidyverse)
library(Rarr)
library(stars)
library(terra)
```

#### Fixes to the `Rarr` package

We will use functions from the `Rarr` package to read and analyse Zarr data. Unfortunately, there is currently a bug in this package, causing it to parse the EOPF Sample Service data URLs incorrectly -- it has been fixed and will be updated in the next release of `Rarr`. In the meantime, we will write our own version of this URL parsing function and use it instead of the one in `Rarr`.

```{r overwrite-url-parsing}
#| cache: false
.url_parse_other <- function(url) {
  parsed_url <- httr::parse_url(url)
  bucket <- gsub(
    x = parsed_url$path, pattern = "^/?([[a-z0-9\\:\\.-]*)/.*",
    replacement = "\\1", ignore.case = TRUE
  )
  object <- gsub(
    x = parsed_url$path, pattern = "^/?([a-z0-9\\:\\.-]*)/(.*)",
    replacement = "\\2", ignore.case = TRUE
  )
  hostname <- paste0(parsed_url$scheme, "://", parsed_url$hostname)

  if (!is.null(parsed_url$port)) {
    hostname <- paste0(hostname, ":", parsed_url$port)
  }

  res <- list(
    bucket = bucket,
    object = object,
    region = "auto",
    hostname = hostname
  )
  return(res)
}

assignInNamespace(".url_parse_other", .url_parse_other, ns = "Rarr")
```

This function overwrites the existing one in `Rarr`, and allows us to continue with the analysis.

If you try to run some of the examples below and receive a timeout error, please ensure that you have run the above code block.

## Access Zarr data from the STAC Catalog

The first step of accessing Zarr data is to understand the assets within the EOPF Sample Service STAC catalog. The [first tutorial](https://eopf-toolkit.github.io/eopf-101/51_eopf_stac_r.html) goes into detail on this, so we recommend reviewing it if you have not already.

For the first part of this tutorial, we will be using data from the [Sentinel-2 Level-2A Collection](https://stac.browser.user.eopf.eodc.eu/collections/sentinel-2-l2a). We fetch the "product" asset under a given item, and can look at its URL:

```{r access-stac-product}
s2_l2a_item <- stac("https://stac.core.eopf.eodc.eu/") |>
  collections(collection_id = "sentinel-2-l2a") |>
  items(feature_id = "S2B_MSIL2A_20250530T101559_N0511_R065_T32TPT_20250530T130924") |>
  get_request()

s2_l2a_product <- s2_l2a_item |>
  assets_select(asset_names = "product")

s2_l2a_product_url <- s2_l2a_product |>
  assets_url()

s2_l2a_product_url
```

The product is the "top level" Zarr asset, which contains the full Zarr product hierarchy. We can use `zarr_overview()` to get an overview of it, setting `as_data_frame` to `TRUE` so that we can see the entries in a data frame instead of printed directly to the console. Each entry is a Zarr array; we remove `product_url` from the path to get a better idea of what each array is. Since this is something we will want to do multiple times throughout the tutorial, we create a helper function for this.

```{r overview}
derive_store_array <- function(store, product_url) {
  store |>
    mutate(array = str_remove(path, product_url)) |>
    relocate(array, .before = path)
}

zarr_store <- s2_l2a_product_url |>
  zarr_overview(as_data_frame = TRUE) |>
  derive_store_array(s2_l2a_product_url)

zarr_store
```

This shows us the path to access the Zarr array, the number of chunks it contains, the type of data, as well as its dimensions and chunking structure.

We can also look at overviews of individual arrays. First, let's narrow down to measurements taken at 20-metre resolution:

```{r measurements-20m}
r20m <- zarr_store |>
  filter(str_starts(array, "/measurements/reflectance/r20m/"))

r20m
```

Then, we select the B02 array and examine its dimensions and chunking:

```{r array-info}
r20m |>
  filter(str_ends(array, "b02")) |>
  select(path, nchunks, dim, chunk_dim) |>
  as.list()
```


We can also see an overview of individual arrays using `zarr_overview()`. With the default setting (where `as_data_frame` is `FALSE`), this prints information on the array directly to the console, in a more digestible way:

```{r array-overview}
r20m_b02 <- r20m |>
  filter(str_ends(array, "b02")) |>
  pull(path)

r20m_b02 |>
  zarr_overview()
```

The above overview tells us that the data is two-dimensional, with dimensions 5490 x 5490. Zarr data is split up into **chunks**, which are smaller, independent piece of the larger array. Chunks can be accessed individually, without loading the entire array. In this case, there are 36 chunks in total, with 6 along each of the dimensions, each of size 915 x 915.

# Read Zarr data

To read in Zarr data, we use `read_zarr_array()`, and can pass a list to the `index` argument, describing which elements we want to extract along each dimension. Since this array is two-dimensional, we can think of the dimensions as rows and columns of the data. For example, to select the first 10 rows and the first 5 columns:

```{r read-array-index}
r20m_b02 |>
  read_zarr_array(index = list(1:10, 1:5))
```

### Coordinates

Similarly, we can read in the `x` and `y` coordinates corresponding to data at 10m resolution. These `x` and `y` coordinates do not correspond to latitude and longitude---to understand the coordinate reference system used in each data set, we access the `proj:code` property of the STAC item. In this case, the coordinate reference system is [EPSG:32626](https://epsg.io/32626), which represents metres from the UTM zone's origin.

```{r item-proj}
s2_l2a_item[["properties"]][["proj:code"]]
```

We can see that `x` and `y` are one dimensional:

```{r x-y-dims}
r20m_x <- r20m |>
  filter(str_ends(array, "x")) |>
  pull(path)

r20m_x |>
  zarr_overview()

r20m_y <- r20m |>
  filter(str_ends(array, "y")) |>
  pull(path)

r20m_y |>
  zarr_overview()
```

Which means that, when combined, they form a grid that describes the location of each point in the 2-dimensional measurements, such as B02. We will go into this more in the examples below.

The `x` and `y` dimensions can be read in using the same logic: by describing which elements we want to extract. Since there is only one dimension, we only need to supply one entry in the indexing list:

```{r x-read-index}
r20m_x |>
  read_zarr_array(list(1:5))
```

Or, we can read in the whole array (by not providing any elements to `index`) and view its first few values with `head()`. Of course, reading in the whole array, rather than a small section of it, will take longer.

```{r x-y-read-head}
r20m_x |>
  read_zarr_array() |>
  head(5)

r20m_y |>
  read_zarr_array() |>
  head(5)
```

### Different resolutions

With EOPF data, some measurements are available at multiple resolutions. For example, we can see that the B02 spectral band is available at 10m, 20m, and 60m resolution:

```{r b02-resolutions}
b02 <- zarr_store |>
  filter(str_starts(array, "/measurements/reflectance"), str_ends(array, "b02"))

b02 |>
  select(array)
```

The resolution affects the dimensions of the data: when measurements are taken at a higher resolution, there will be more data. We can see here that there is more data for the 10m resolution than the 20m resolution (recall, its dimensions are 5490 x 5490), and less for the 60m resolution:

```{r b02-resolutions-dims}
b02 |>
  filter(array == "/measurements/reflectance/r10m/b02") |>
  pull(dim)

b02 |>
  filter(array == "/measurements/reflectance/r60m/b02") |>
  pull(dim)
```
